Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-09-16 submission/manuscript.Rmd
adjust labels and figure legend

\begin{figure}
  \centering
  \includegraphics{../submission/compiled_figs/KDM5C_Mechanism.pdf}
  \caption[KDM5C promotes long-term silencing of germline genes via DNA methylation at CpG islands]{\textbf{KDM5C promotes long-term silencing of germline genes via DNA methylation at CpG islands.} \textbf{A.} Diagram of embryonic stem cell (ESC) to epiblast-like cell (EpiLC) differentiation and collection time points. \textbf{B.} Real time quantitative PCR (RT-qPCR) of \textit{Kdm5c} mRNA expression in wild-type (WT) and \textit{Kdm5c}-KO (5CKO) ESCs at 0, 24, and 48 hours of differentiation into EpiLCs. Expression calculated in comparison to TBP mRNA expression (2\textsuperscript{-deltaCT}). * p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test. \textbf{C.} KDM5C protein expression normalized to DAXX. Quantified intensity using ImageJ (artificial units - au). Right - representative lanes of Western blot for KDM5C and DAXX. * p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test. \textbf{D.} Top - Representative UCSC browser view of histone 3 lysine 4 dimethylation (H3K4me2) ChIP-seq signal at two germline genes in wild-type and \textit{Kdm5c}-KO EpiLCs. Bottom - Average H3K4me2 signal at the TSS of all germline-enriched genes in wild-type (dark purple) and \textit{Kdm5c}-KO (light purple) EpiLCs. \textbf{E.} Top - Representative UCSC browser view of histone 3 lysine 4 trimethylation (H3K4me3) ChIP-seq signal at two germline genes in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) adult amygdala. Bottom - Average H3K4me3 signal at the transcription start site (TSS) of all germline-enriched genes in wild-type (dark red) and \textit{Kdm5c}-KO (light red) amygdala. \textbf{F.} Percentage of germline genes that harbor CpG islands (CGIs) in their promoters (CGI promoter, red), based on UCSC annotation. Left - percentage of all germline-enriched genes, middle - KDM5C-bound germline genes, and right - KDM5C-unbound germline genes. (Legend continued on next page.)}
  \label{figurelabel}
\end{figure}
\newpage
\addtocounter{figure}{-1}
\begin{figure} [t!]
  \caption[]{\textbf{KDM5C promotes long-term silencing of germline genes via DNA methylation at CpG islands.} (Legend continued.) \textbf{G.} enrichPlot gene ontology analysis of germline genes with (CGI-promoter) or without (CGI-free) CGIs in their promoter. \textbf{H.} Expression of germline genes with (CGI-promoter, red) or without (CGI-free, salmon) CGIs in their promoter across stages of spermatogenesis from Green et al 2018. SPG - spermatogonia, Pre-Lep - preleptotene spermatocytes, S-Cytes - meiotic spermatocytes, S-Tids - post-meiotic haploid round spermatids, E.Sp - elongating spermatids. Wilcoxon test, * p < 0.05, ** p < 0.01, *** p < 0.001, **** p < 0.0001. \textbf{I.} Diagram of ESC to extended EpiLC (exEpiLC) differentiation. \textbf{J.} Volcano plot of whole genome bisulfite sequencing (WGBS) comparing CpG methylation (CpGme) at germline gene promoters (TSS $\pm$ 500) in wild-type (WT) ESCs versus exEpiLCs. Significantly differentially methylated promoters (q < 0.01, |methylation difference| > 25\%) with CGIs in red, CGI-free promoters in light blue \textbf{K.} Volcano plot of WGBS of wild-type (WT) versus \textit{Kdm5c}-KO exEpiLCs for germline gene promoters. Promoters with CGIs in red, CGI-free promoters in light blue. \textbf{L.} Histogram of average percent CpGme at the promoter of germline genes with or without CGIs. Wild-type in navy and \textit{Kdm5c}-KO (KO) in light blue. Dashed lines are average methylation for each genotype, p-values for Mann-Whitney U test. \textbf{M.} UCSC browser view of around the transcription start site (TSS) of germline genes, showing UCSC annotated CGI, representative CpGme in wild-type (WT) and \textit{Kdm5c}-KO (5CKO) ESCs and exEpiLCs.}
\end{figure}

>>>>

\begin{figure}
  \centering
  \includegraphics{../submission/compiled_figs/KDM5C_Mechanism.pdf}
  \caption[KDM5C promotes long-term silencing of germline genes via DNA methylation at CpG islands]{\textbf{KDM5C promotes long-term silencing of germline genes via DNA methylation at CpG islands.} \textbf{A.} Diagram of embryonic stem cell (ESC) to epiblast-like cell (EpiLC) differentiation and collection time points. \textbf{B.} Real time quantitative PCR (RT-qPCR) of \textit{Kdm5c} mRNA expression in wild-type (WT) and \textit{Kdm5c}-KO (5CKO) ESCs at 0, 24, and 48 hours of differentiation into EpiLCs. Expression calculated in comparison to TBP mRNA expression (2\textsuperscript{-deltaCT}). * p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test. \textbf{C.} KDM5C protein expression normalized to DAXX. Quantified intensity using ImageJ (artificial units - au). Right - representative lanes of Western blot for KDM5C and DAXX. * p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test. \textbf{D.} Top - Representative UCSC browser view of histone 3 lysine 4 dimethylation (H3K4me2) ChIP-seq signal at two germline genes in wild-type and \textit{Kdm5c}-KO EpiLCs. Bottom - Average H3K4me2 signal at the TSS of all germline-enriched genes in wild-type (dark purple) and \textit{Kdm5c}-KO (light purple) EpiLCs. \textbf{E.} Top - Representative UCSC browser view of histone 3 lysine 4 trimethylation (H3K4me3) ChIP-seq signal at two germline genes in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) adult amygdala. Bottom - Average H3K4me3 signal at the transcription start site (TSS) of all germline-enriched genes in wild-type (dark red) and \textit{Kdm5c}-KO (light red) amygdala. \textbf{F.} Percentage of germline genes that harbor CpG islands (CGIs) in their promoters (CGI promoter, red), based on UCSC annotation. Left - percentage of all germline-enriched genes, middle - KDM5C-bound germline genes, and right - KDM5C-unbound germline genes. (Legend continued on next page.)}
  \label{figurelabel}
\end{figure}
\newpage
\addtocounter{figure}{-1}
\begin{figure} [t!]
  \caption[]{\textbf{KDM5C promotes long-term silencing of germline genes via DNA methylation at CpG islands.} (Legend continued.) \textbf{G.} enrichPlot gene ontology analysis of germline genes with (CGI-promoter) or without (CGI-free) CGIs in their promoter. \textbf{H.} Expression of germline genes with (CGI-promoter, red) or without (CGI-free, salmon) CGIs in their promoter across stages of spermatogenesis from Green et al 2018. SPG - spermatogonia, Pre-Lep - preleptotene spermatocytes, S-Cytes - meiotic spermatocytes, S-Tids - post-meiotic haploid round spermatids, E.Sp - elongating spermatids. Wilcoxon test, * p < 0.05, ** p < 0.01, *** p < 0.001, **** p < 0.0001. \textbf{I.} Diagram of ESC to extended EpiLC (exEpiLC) differentiation. \textbf{J.} Volcano plot of whole genome bisulfite sequencing (WGBS) comparing CpG methylation (CpGme) at germline gene promoters (TSS $\pm$ 500) in wild-type (WT) ESCs versus exEpiLCs. Significantly differentially methylated promoters (q < 0.01, |methylation difference| > 25\%) with CGIs in red, CGI-free promoters in light blue \textbf{K.} Volcano plot of WGBS of wild-type (WT) versus \textit{Kdm5c}-KO exEpiLCs for germline gene promoters. Promoters with CGIs in red, CGI-free promoters in light blue. \textbf{L.} Histogram of average percent CpGme at the promoter of germline genes with or without CGIs. Wild-type in navy and \textit{Kdm5c}-KO (KO) in light blue. Dashed lines are average methylation for each genotype, p-values for Mann-Whitney U test. \textbf{M.} UCSC browser view of around the transcription start site (TSS) of germline genes, showing UCSC annotated CGI with number of CpGs, representative CpGme in wild-type (WT) and \textit{Kdm5c}-KO (5CKO) ESCs and exEpiLCs, and KDM5C ChIP-seq in EpiLCs.}
\end{figure}
